Solving Variational Inequalities and Cone Complementarity Problems in Non-Smooth Dynamics using the Alternating Direction Method of Multipliers

### 2021-02-16 2021/ADMM_journal/author.tex
Rephrase
Added changes requested by reviewers after revision 1. Notation section, better bibliography, corrected typos, added some sentences.

\documentclass[final,3p]{elsarticle}

% to typeset URLs, URIs, and DOIs
\usepackage{url}
%\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{cleveref}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage[ruled,vlined]{algorithm2e}  % ruled,vlined,linesnumbered
\usepackage{array}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage[title]{appendix}
\usepackage{empheq}

@@

\newtheorem*{definition*}{Definition}
\newtheorem{definition}{Definition}
\newtheorem*{remark*}{Remark}
\newtheorem{remark}{Remark}
\newtheorem*{theorem*}{Theorem}
\newtheorem{theorem}{Theorem}[section]
\newtheorem*{corollary*}{Corollary}
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

@@

\title{Solving Cone Complementarity Problems in Non-Smooth Dynamics using the Alternating Direction Method of Multipliers}

%\author{Alessandro Tasora \\ alessandro.tasora@unipr.it}
\author{Alessandro Tasora}
\ead{alessandro.tasora@unipr.it}
\author{Dario Mangoni}
\ead{dario.mangoni@unipr.it}
\author{Simone Benatti}
\ead{simone.benatti@studenti.unipr.it}
\author{Rinaldo Garziera}
\ead{rinaldo.garziera@unipr.it}
\address{Universit\`a degli Studi di Parma, \\Dpt. of Engineering and Architecture\\Parco Area delle Scienze, 181/A, 43124 Parma, Italy }

@@

\begin{abstract}
We propose to use the Alternating Direction Method of Multipliers (ADMM) for solving the variational inequality that arises in non-smooth contact problems in multibody simulations. The ADMM method is based on simple computational primitives and offers good convergence properties especially for scenarios where loose tolerance on the precision can be accepted, thus making it attractive also for real-time applications. The method scales well with the problem size and require a small number of tuning parameters. Differently from other methods known in literature, such as fixed point iterations, it can easily handle problems that mix finite elements and rigid bodies, and it is only marginally affected by odd mass ratios or ill-posed problems. Another attractive feature is that it can be easily warm-started.
\end{abstract}
%
\begin{keyword}
ADMM \sep non-smooth dynamics \sep contact
\end{keyword}

@@

Following the seminal works by Moreau 
\cite{mor88,Jean1992} 
most formulations for non-smooth dynamical problems are based on Measure Differential Inclusions (MDI): just like differential inclusions they allow set-valued force laws (such as the Coulomb-Amontons dry friction model) but also generalize to the case where velocity is assumed to be a possibly discontinuous function of bounded variation in order to allow impulsive events.

These problems can be solved by means of special time stepping methods that offer superior robustness and stability at the cost of solving a complementarity problem, or more in general a Variational Inequality (VI), per each time step
 \cite{acary2008numerical}. 
In this context, unknowns to be solved are velocity measures and reaction impulses at contact points and at joints: in cases of many parts with lot of frictional contacts, the large dimension of the VI could lead to a bottleneck in the time stepping process. This stimulated lot of research on efficient numerical methods in the last three decades.

One of the former approaches, presented in 
\cite{StTr95}  % LCP solved via Lemke
, was based on solving a Linear Complementarity Problem (LCP) per each time step. LCPs are sub cases of VIs, for whom a direct method (Lemke algorithm) does exist. However, direct LCP solvers offer exact solution at the cost of very expensive pivoting sub iterations that scale badly with increasing number of unknowns, and for these reason they are not much used nowadays.

On the other hand, approximate but efficient methods based on fixed-point iterations became popular in the area of real-time simulators and robotics \cite{Bender2014}.
In most cases they are based on stationary methods like Gauss-Seidell or Jacobi iterations interleaved with projections on friction cones. They can be used to solve the LCP or the Cone Complementarity Problem (CCP), another special case of the VI in which the multibody contact problem can be formulated.
Attempts were made in order to parallelize them and to increase their convergence, such as over-relaxation, Krasnoselskii-Mann smoothing and warm starting, 
\cite{massSplittingRichard2012,TasoraAnitescuCMAME10} % P-GS iteration
but in general they perform poorly when there are odd mass ratios, and when articulated mechanisms such as robots are added to the scenarios - in those cases convergence often stalls, and if the iteration is prematurely truncated, reaction impulses are badly estimated and mechanisms fall apart or bend, and objects might interpenetrate.

Another option is to solve the CCP as an optimization problem, using first-order optimization methods such as 
the Nesterov accelerated projected gradient descend  
\cite{hammadTOG2015}
or the Barzilai-Borwein spectral projected gradient
\cite{heynIJNME2013}. These methods are based on a simple projection operator,  matrix-by-vector multiplications and inner products. Their convergence is better than for fixed points iterations, however they share with them the following issue: that they often require the building of a Delassus operator, a matrix that is easy to compute in efficient (factored and sparse) format if diagonal masses are used because these can be easily inverted, but hard to handle if finite elements are added to the problem, because they introduce stiffness and damping matrices whose inverse would be needed as well.

A further class of solvers is represented by non-smooth newton methods, such as the one presented in 
\cite{Macklin2019} % Non-smooth Newton Method
, that assume a generic Nonlinear Complemetarity Problem (NCP), again a sub case of a VI, and represent the NCP complementarity constraints using non-smooth functions like the Fisher-Burmeister function. A generalized non-smooth Newton method can be used to find the zero of the functions, at the cost of solving a linear system per each iteration.

Under mild assumptions, the VI can be stated as a convex CCP, that can be also cast as an optimization problem, namely a Quadratic Program (QP) with convex constraints \cite{anitescuTasora2008}. For such a class of optimization problems, one can use Interior Point Methods (IPMs), a class of solvers that share some similarities with the above mentioned non-smooth Newton methods - for example the solution of a saddle-point linear system is required at each iteration. IPMs offer the best theoretical convergence. However, their implementation is quite intricate, and despite the encouraging theoretical properties, in practice they do not scale well for large problems. Also, there are no reliable ways to warm-start them
\cite{Mangoni2018}
.

The Alternating Direction Method of Multipliers (ADMM) has been proposed in the 1970s 
\cite{Glowinski75} % [ ADMM invention ]
\cite{Gabay1976} % [ ADMM invention ]
as a practical way to solve constrained optimization problems by iterating over the minimization of two (not necessarily differentiable) objectives.
ADMM introduces auxiliary variables and cast the original problem as the minimization of a separable objective subject to a linear constraint between the original variables and the auxiliary variables. With a proper choice of auxiliary variables, the minimization of the separable function is obtained iterating on two distinct minimization steps of lower complexity. The method iterates updating primal variables, auxiliary variables and dual variables similarly to a fixed point iteration: as such, they compares less favorably to IPMs if one is interested in convergence to high precision, however recent developments showed that in practical scenarios they offer superior speed, scalability and robustness, moreover their structure map well into parallel computing architectures
 \cite{Boyd2011}. % [ ADMM general theory and development ].
These positive properties triggered a recent revival of ADMM and similar operator-splitting methods even as alternatives to IPMs, especially in distributed convex optimization with large-scale problems where moderate precision is acceptable
\cite{Cannon2019}. % [COSMO  conic operator splitting method for large convex problems].

ADMM methods have been used in many fields, for instance computer graphics 
 \cite{Zhang2019}
, computational fluid dynamics
 \cite{Gregson2014} % [ ADMM and fluid simulation, inverse problems ]
, simulation of deformable structures 
 \cite{Overby2017} % [ ADMM projective dynamics hyperelastic models ]
and contact dynamics 
 \cite{Daviet2020} % [ ADMM thin objects frictional contact ]
, to name a few.

Recent developments addressed the possibility of accelerating the ADMM method.
In 
 \cite{Goldfarb2013} % [ ADMM with acceleration, both convex (non-strongly-convex require step skipping), one differentiable]
an acceleration scheme has been proposed for special cases of symmetric ADMMs, and assuming that one of the two objectives is differentiable.
%
In 
 \cite{Goldstein2014} % [ ADMM and AMA with Nesterov. Convergence ok if both problems are strongly convex (non-strongly-convex require restarting), one is quadratic ].
the Nesterov acceleration method has been proposed for problems where at least one of the two objectives is quadratic, and under the assumption that both objectives are strongly convex; if the latter assumption is not guaranteed, restarting strategies are needed. 
%
In 
 \cite{Kadkhodaie2015} %  [ A2DM2 accelerated ADMM, strongly convex ]
a similar acceleration method has been proposed, with $\mathcal{O}(\frac{1}{k^2})$ convergence rate, but without the assumption that one of the objectives must be quadratic.
%
Other types of improvements, based on the Anderson fixed point acceleration, have been put forward in 
 \cite{Zhang2019} % [ ADMM + Anderson acceleration, computergraphics ].
 \cite{Ouyang2020} % [ ADMM + Anderson acceleration, using Douglas-Rachford splitting, computergraphics ]
, showing high performance in computer graphics applications.

Recent developments also focused on the possibility of parallelizing the ADMM iterations on GPU and parallel architectures
\cite{Schubiger2020} % [ GPU implementation of ADMM | no need to compute residuals each iteration | krylov solver each iteration ].
.

In
\cite{Stellato2020} % [OSQP: an operator splitting solver for quadratic programs].
a specialized ADMM method for quadratic problems with conic constraints has been presented: on the top of this method we develop a solver that can exploit the nature of non-smooth dynamical problems. An attractive property of such ADMM is that it requires the solution of a linear system per each iteration, but unlike IPMs, such linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.

In our embodiment, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), the convex cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).

@@

@@

\begin{definition}
The \textit{projection} of $\vect{y}$ on a nonempty closed convex set $\mathcal{E}$ is a $\mathbb{R}^n \rightarrow \mathbb{R}^n$ mapping defined as:
%
\begin{align}
\label{eq:projector}
\Pi_{\mathcal{E}}(\vect{y}) = \mathrm{argmin}_{\vect{x} \in \mathcal{E}} \norm{ \vect{x} - \vect{y} }^2_2
\end{align}
%
If $f(\vect{y})$ is the indicator function $I_{\mathcal{E}}(\vect{y})$ to a nonempty closed convex set ${\mathcal{E}}$, defined as in \eqref{eq:indicator}, then $\mathrm{prox}_{f}(\vect{y})$ is the orthogonal projection onto that set:
%
\begin{align}
\label{eq:projectorindicator}
\mathrm{prox}_{I_{\mathcal{E}}}(\vect{y})  = \Pi_{\mathcal{E}}(\vect{y}) 
\end{align}
\label{def:projection}
\end{definition}

@@

where 
\begin{itemize}
		\item unknown $\vect{v} \in \mathbb{R}^{n_v}$ is the speed at the end of the time step $h$
		\item unknown $\vect{\gamma}_\epsilon \in \mathbb{R}^{n_c}$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions)
    \item $H \in \mathbb{R}^{n_v \times n_v}$ is a positive definite matrix containing $M$, 
		the block-diagonal matrix of the masses and inertia tensors of the bodies; if stiff
		loads are added too (ex. when finite elements come into play) it becomes a block-sparse matrix including the tangent stiffness and damping matrices,
		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla_v\vect{f}$ in case of the first-order time stepper in \cite{TasoraAnitescuCMAME10}
    \item $D_\epsilon \in \mathbb{R}^{n_v \times n_c}$ is a sparse matrix, the transpose jacobian of all constraints
    \item $\vect{k} \in \mathbb{R}^{n_v}$ is a vector containing terms proportional to applied forces $\vect{f}$ and to the last known speed,
    \item $\vect{b} \in \mathbb{R}^{n_c}$ is a vector containing constraint stabilization terms and rheonomic terms,
    \item $\Upsilon$ is the Cartesian product of all cones of admissible constraint forces, $\Upsilon = \bigtimes_j \Upsilon_j$
    \item if a frictional contact is added, $\Upsilon_j \subset \mathbb{R}^3$ is a second order Lorentz cone with aperture proportional to friction coefficient
    \item if a bilateral constraint is added, $\Upsilon_j = \mathbb{R}$ and $\Upsilon^\circ_j = \{0\}$
    \item if a unilateral constraint is added, $\Upsilon_j = \mathbb{R}^+$ and $\Upsilon^\circ_j = \mathbb{R}^-$
    \item $\Upsilon^{\circ}$ is the polar cone, opposite of the dual cone, i.e. $\Upsilon^{*} = -\Upsilon^\circ$.
\end{itemize}

@@

This CCP corresponds exactly to a first order optimality condition of a convex quadratic program:
\begin{subequations}
	\begin{empheq}[box=\fbox]{align}
	\text{min} \quad & \frac{1}{2} \vect{\gamma}_\epsilon^T N \vect{\gamma}_\epsilon + \vect{r}^T_\epsilon \vect{\gamma}_\epsilon \\
	\text{s.t.} \quad & \vect{\gamma}_\epsilon \in \Upsilon
	\end{empheq}
	\label{eq:ChronoCCP_min}
\end{subequations}
%
and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the KKT optimality conditions on \textit{dual variables} $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$, and \textit{primal variables} $\vect{\gamma}_\epsilon$:
\begin{subequations}
	\begin{align}
    N \vect{\gamma}_\epsilon + \vect{r}_\epsilon + I \vect{y} &= \vect{0} \\
    \vect{\gamma}_\epsilon &= \vect{z} \\
    \Upsilon \ni \vect{z}  \quad \bot &\quad \vect{y} \in \Upsilon^\circ  
	\end{align}
	\label{eq:ChronoCCP_kkt}
\end{subequations}
After the convex program \eqref{eq:ChronoCCP_min} is solved by ADMM, one can compute 
\begin{align}
\vect{v} = H^{-1}( \vect{k} + D_\epsilon \vect{\gamma}_\epsilon)
\label{eq:v_post}
\end{align}
with an immediate step.

@@

This is an unconstrained convex quadratic program whose KKT optimality conditions lead to the following linear problem:
\begin{align}
    \left[ N + \rho I \right] \vect{\gamma}_\epsilon^{k+1} =  \rho \vect{z}^k - \vect{r} - \vect{y}^k
		\label{eq:admm_1b}
\end{align}

@@

We experienced that it is possible to use a $\rho$ value that changes on a per-constraint basis. If one could know in advance which contact will be active at the solution, one could split $\rho_i$ values between very high or very low values for the fastest convergence, however this is not possible because the set of active constraints is known only at the solution. However, we still can use a
simple heuristics consisting in forcing a small value $\rho_b = \SI{1e-9}{}$ for bilateral constraints, where one knows in advance that the corresponding dual variable $y_i$ would be null anyway. This produces better convergence results in systems containing both contacts and many bilateral joints, such as when simulating articulated robots that grasp some objects.

>>>>

\documentclass[AMA,STIX1COL]{WileyNJD-v2}

\articletype{Article Type}%

\received{28 December 2020}
\revised{- - -}
\accepted{- - -}

\raggedbottom

% to typeset URLs, URIs, and DOIs
\usepackage{url}
%\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{cleveref}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{siunitx}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{tabularx}
%\usepackage[ruled,vlined]{algorithm2e}  % ruled,vlined,linesnumbered % NOT WORKIGN IN IJNME, they use algorithmicx
\usepackage{array}
\usepackage{booktabs}
\usepackage{multicol}
\usepackage{xcolor}
%\usepackage[title]{appendix}
\usepackage{empheq}
\usepackage{framed} % Framing content
\usepackage{nomencl} % Nomenclature package
\makenomenclature
\setlength{\nomlabelwidth}{2cm}
\setlength{\nomitemsep}{-0.1cm} % Baseline skip between items

@@

\newtheorem*{definition*}{Definition}
%\newtheorem{definition}{Definition}
\newtheorem*{remark*}{Remark}
%\newtheorem{remark}{Remark}
\newtheorem*{theorem*}{Theorem}
%\newtheorem{theorem}{Theorem}[section]
\newtheorem*{corollary*}{Corollary}
%\newtheorem{corollary}{Corollary}[theorem]
%\newtheorem{lemma}[theorem]{Lemma}

@@

\title{Solving Variational Inequalities and Cone Complementarity Problems in Non-Smooth Dynamics using the Alternating Direction Method of Multipliers}

%\author{Alessandro Tasora \\ alessandro.tasora@unipr.it}
\author{Alessandro Tasora*}

\author{Dario Mangoni}

\author{Simone Benatti}

\author{Rinaldo Garziera}

\address{\orgdiv{Dpt. of Engineering and Architecture}, \orgname{University of Parma}, \orgaddress{Parco Area delle Scienze, 181/A, 43124 Parma, \country{Italy}} }

\corres{*Corresponding Author. \email{alessandro.tasora@unipr.it}}

@@

\abstract[Summary]{
This work presents a numerical method for the solution of variational inequalities arising in non-smooth flexible multi-body problems that involve set-valued forces. For the special case of hard frictional contacts, the method solves a second order cone complementarity problem. We ground our algorithm on the Alternating Direction Method of Multipliers (ADMM), an efficient and robust optimization method that draws on few computational primitives. In order to improve computational performance, we reformulated the original ADMM scheme in order to exploit the sparsity of constraint jacobians and we added optimizations such as warm starting and adaptive step scaling. The proposed method can be used in scenarios that pose major difficulties to other methods available in literature for complementarity in contact dynamics, namely when using very stiff finite elements and when simulating articulated mechanisms with odd mass ratios. The method can have applications in the fields of robotics, vehicle dynamics, virtual reality, and multi-physics simulation in general.
}

\keywords{
Contact, friction, ADMM, non-smooth dynamics, simulation
}

@@

Problems involving frictional contacts can be found in many engineering fields. Their simulation is a challenging task because the discontinuous nature of contact phenomena discourages the adoption of conventional time integration schemes, which assume continuous velocities and accelerations. In fact, if on the one hand it is possible to turn a discontinuous model into a smooth Ordinary Differential Equation (ODE) by approximating contact forces via spring-damper penalty forces, on the other this usually requires the adoption of extremely small time steps.

In search of better numerical performance and stability, even in case of large time steps, we rather formulate the original non-smooth dynamical problem as a Measure Differential Inclusion (MDI). MDIs were pioneered by Moreau in the eighties, among others, aiming at a viable approach to the simulation of mechanical systems with hard contacts
\cite{Moreau1983,moreau1988topics}. 
Being differential inclusions, they can directly embed set-valued force laws, such as the Coulomb friction model, but they also generalize to the case where velocity is assumed to be a (possibly) discontinuous function of bounded variation, therefore they admit impulsive events.

MDIs can be solved using special time stepping methods that offer high stability and robustness, however this comes at the cost of solving a Complementarity Problems (CP) per each time step \cite{acary2008numerical}. These CPs correspond to the broader class of Variational Inequalities (VI) \cite{Kinderleher1980,Harker1990}.

In general, the solution of VIs or CPs represents the major numerical bottleneck in the entire time stepping process of the MDI, especially when dealing with many parts or many contacts. For instance, similar scenarios can happen when simulating packaging devices, ground-machine interaction, rock dynamics, masonry structures and granular materials, where the unknowns of the VIs (contact forces) can amount to thousands or millions. This challenge stimulated research on efficient numerical methods in the last three decades.

Among the former strategies for solving this class of problems, we cite the idea of solving a Linear Complementarity Problem (LCP) per each time step
\cite{StTr95}  % LCP solved via Lemke
.
Once cast as a LCP, the VI problem can be solved by direct LCP solvers such as the Lemke algorithm. However, LCP direct solvers often lead to complex sub-iterations that do not scale well with increasing number of unknowns, hence they are not much used nowadays. Another drawback of casting the problem as a LCP is that they introduce a polyhedral approximation, for example the Coulomb friction cone would be approximated as a faceted pyramid, introducing a numerical source of anisotropy.

At the other end of the spectrum are iterative methods that aim at high robustness and efficiency even at the expense of sacrificing accuracy, since this is often acceptable in fields like real time simulation, video games and virtual reality \cite{Bender2014}.
In these fields, the most used approach nowadays consists in fixed point iterations, similarly to Gauss-Seidel, SOR or Jacobi stationary methods, interleaved with simple projections onto the constraint sets at each iteration
\cite{anitescuTasora2008,massSplittingRichard2012,AleMihaiFriction2013} % P-GS iteration
.
The major issue of these methods is that they exhibit slow convergence when the system contains long kinematic chains or long sequences of contacts, such as when including high stacks of objects \cite{TasoraAnitescuCMAME10}. Their convergence is even more challenged if odd mass ratios are encountered: this often happens when simulating industrial robots, earth-moving machines, complex vehicle suspensions, car drivelines. In these cases the convergence might stall soon, hence if the iterations are truncated in order to meet a time budget in real-time applications, the resulting simulation can be affected by low accuracy - usually detectable in the form of objects that interpenetrate at the point of contact and mechanisms that fall apart. Optimizations like warm starting and sub-stepping can alleviate the drawbacks of these methods \cite{Macklin2019a}.

A more advanced class of solvers is represented by non-smooth Newton methods, such as the one presented in 
\cite{Macklin2019} % Non-smooth Newton Method
, that assume a generic Nonlinear Complemetarity Problem (NCP), again a sub case of a VI, and that enforce the NCP complementarity constraints using non-smooth functions like the Fisher-Burmeister function. In these methods, a generalized non-smooth Newton method can be used to find the zero of the functions, but this comes at the cost of solving a linear system per each iteration.

Recent efforts pushed forward the idea of casting the VI of non-smooth dynamics as a convex optimization problem, namely a Quadratic Program (QP) with convex constraints. This requires a simplifying assumption about the Coulomb friction being associative instead of non-associative. This artificial associativity of the friction model means that the relative sliding velocity in a contact point will be restricted to the dual cone of the Coulomb friction cone, whereas the original model has a more relaxed condition, assuming that such velocity could be even parallel to the surface: the practical consequence is that grazing motion cause an artificial lift-off - however this artifact can be attenuated thanks to the introduction of a stabilization term \cite{ani04}. In this setting, the problem is also a convex Cone Complementarity Problem (CCP). The convexity assumption allows the adoption of numerical methods already available for large scale optimization problems. In \cite{heynIJNME2013}, the Barzilai-Borwein spectral projected gradient is used to this end, and in \cite{hammadTOG2015} we used the Nesterov accelerated projected gradient descend to simulate large scale contact problems.
These are first-order optimization methods, in the sense that they do not require the computation of the Hessian matrix: in fact they rely only on  matrix-by-vector multiplications and inner products (similarly to Krylov subspace methods in the field of linear systems), plus a fast projection operator. Although they offer superior convergence with respect to fixed point iterations, they are not easily applicable to problems involving finite elements. In fact, this class of solvers often perform the matrix-by-vector primitive over a Schur complement matrix, a product of constraint jacobians and inverse of the mass matrix. Since the mass matrix is often (block) diagonal, the Schur complement can be quickly factored at the beginning of the iteration, but if finite elements are used, the Schur matrix would also involve to some extent the inverse of element stiffness matrices and damping matrices, a costly and inefficient operation unless special developments are considered \cite{Francu2017}.
Extending complementarity-based approaches in order to encompass finite elements is important because recent research has shown that complementarity-based formulations for large deformation dynamics with contacts often show superior performance when compared to penalty-based approaches \cite{Yu2020,Bozorgmehri2021}.

Another class of solvers that can draw on the optimization-based approach are Interior Point Methods (IPMs). These solvers share some similarities with the previously mentioned non-smooth Newton methods, in that they require the solution of a saddle-point linear system at each iteration. IPMs offer the best theoretical convergence for the problem at hand \cite{Potra2000}. However, their implementation is complex, and despite the encouraging theoretical properties, in practical scenarios they do not scale well for large problems. One of their drawbacks, also, is that at the present state of knowledge there are no efficient ways to warm-start them \cite{Mangoni2018}. This led us to investigate other methods for constrained optimization problems.

The Alternating Direction Method of Multipliers (ADMM) has been proposed in the 1970s 
\cite{Glowinski75} % [ ADMM invention ]
\cite{Gabay1976} % [ ADMM invention ]
as a practical way to solve constrained optimization problems by iterating over the minimization of two (not necessarily differentiable) objectives.
ADMM introduces auxiliary variables and casts the original problem as the minimization of a separable objective subject to a linear constraint between the original variables and the auxiliary variables. With a proper choice of auxiliary variables, the minimization of the separable function is obtained iterating on two distinct minimization steps of lower complexity. The method iterates updating primal variables, auxiliary variables and dual variables similarly to a fixed point iteration: as such, they compare less favourably to IPMs if one is interested in convergence to high precision, however recent developments showed that in practical scenarios they offer superior speed, scalability and robustness
 \cite{Boyd2011}. % [ ADMM general theory and development ].
These positive properties triggered a recent revival of ADMM and similar operator-splitting methods even as alternatives to IPMs, especially in distributed convex optimization with large-scale problems where moderate precision is acceptable
\cite{Cannon2019}. % [COSMO  conic operator splitting method for large convex problems].

ADMM methods have been used in many fields, with recent and notable examples in computer graphics 
 \cite{Zhang2019},
computational fluid dynamics
 \cite{Gregson2014}, % [ ADMM and fluid simulation, inverse problems ]
simulation of deformable structures 
 \cite{Overby2017}, % [ ADMM projective dynamics hyperelastic models ]
and contact dynamics 
 \cite{Daviet2020,LeLidec2020}. % [ ADMM thin objects frictional contact ] % [ ADMM in appendix for differentiable dynamics ]

Recent developments addressed the possibility of accelerating the ADMM method.
In 
 \cite{Goldfarb2013} % [ ADMM with acceleration, both convex (non-strongly-convex require step skipping), one differentiable]
an acceleration scheme has been proposed for special cases of symmetric ADMMs, and assuming that one of the two objectives is differentiable.
%
In 
 \cite{Goldstein2014} % [ ADMM and AMA with Nesterov. Convergence ok if both problems are strongly convex (non-strongly-convex require restarting), one is quadratic ].
the Nesterov acceleration method has been proposed for problems where at least one of the two objectives is quadratic, and under the assumption that both objectives are strongly convex; if the latter assumption is not guaranteed, restarting strategies are needed. 
%
In 
 \cite{Kadkhodaie2015} %  [ A2DM2 accelerated ADMM, strongly convex ]
a similar acceleration method has been proposed, with $\mathcal{O}\left(\frac{1}{k^2}\right)$ convergence rate, but without the assumption that one of the objectives must be quadratic.
%
Other types of improvements, based on the Anderson fixed point acceleration, have been put forward in 
 \cite{Zhang2019} % [ ADMM + Anderson acceleration, computergraphics ].
 \cite{Ouyang2020} % [ ADMM + Anderson acceleration, using Douglas-Rachford splitting, computergraphics ]
, showing high performance in computer graphics applications.

Recent developments also focused on the possibility of parallelizing the ADMM iterations on GPU and parallel architectures
\cite{Schubiger2020}. % [ GPU implementation of ADMM | no need to compute residuals each iteration | krylov solver each iteration ].

In
\cite{Stellato2020} % [OSQP: an operator splitting solver for quadratic programs].
a specialized ADMM method for quadratic problems (OSQP) has been presented: building off this method, we develop a solver that can exploit the nature of non-smooth dynamical problems. An attractive property of the approach is that it requires the solution of a linear system per each iteration, but unlike IPMs, the linear system most often remains unchanged during the iterations, so a factorization can be reused multiple times with a great benefit in terms of speed.

In the approach proposed herein, unknowns are primal variables (velocity measures) and dual variables (impulses in contact points and joints), the convex cone constraints stem from the Coulomb friction law, and the system matrix includes terms from the mass matrices and from the tangent stiffness/damping matrices (hence it is sparse, but not necessarily diagonal).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% NOTE FOR THE TYPESETTING DEPARTMENT: IF THE NOMENCLATURE FRAME
%% CONSUMES TOO MUCH SPACE IN THE ARTICLE PAGES, YOU CAN ALSO REMOVE IT. WE ARE
%% PROPOSING IT, BUT WE CAN ALSO DO WITHOUT IT, IF WE ARE GOING OUT OF THE PAGE BUDGET.

    \nomenclature[10]{ADMM}{Alternating Direction Method of Multipliers}
    \nomenclature[10]{FADMM}{Fast Alternating Direction Method of Multipliers}
    \nomenclature[20]{VI}{Variational Inequality}
    \nomenclature[30]{MDI}{Measure Differential Inclusion}
    \nomenclature[40]{CP}{Complementarity Problem}
    \nomenclature[40]{NCP}{Nonlinear Complementarity Problem}
    \nomenclature[50]{LCP}{Linear Complementarity Problem}
    \nomenclature[60]{CCP}{Cone Complementarity Problem}
    \nomenclature[70]{QP}{Quadratic Program}
    %\nomenclature[80]{IPM}{Interior Point Method}

    \begin{table*}[!t]
       \begin{framed}
         \printnomenclature
       \end{framed}
    \end{table*}

@@

\begin{definition}
The \textit{projection} of $\vect{y}$ on a nonempty closed convex set $\mathcal{E}$ is a $\mathbb{R}^n \rightarrow \mathbb{R}^n$ mapping defined as:
%
\begin{align}
\label{eq:projector}
\Pi_{\mathcal{E}}(\vect{y}) = \mathrm{argmin}_{\vect{x} \in \mathcal{E}} \norm{ \vect{x} - \vect{y} }^2_2
\end{align}
\end{definition}

\begin{remark*}
If $f(\vect{y})$ is the indicator function $I_{\mathcal{E}}(\vect{y})$ to a nonempty closed convex set ${\mathcal{E}}$, defined as in \eqref{eq:indicator}, then $\mathrm{prox}_{f}(\vect{y})$ is the orthogonal projection onto that set:
%
\begin{align}
\label{eq:projectorindicator}
\mathrm{prox}_{I_{\mathcal{E}}}(\vect{y})  = \Pi_{\mathcal{E}}(\vect{y}) 
\end{align}
\label{def:projection}
\end{remark*}

@@

where 
\begin{itemize}
		\item unknown $\vect{v} \in \mathbb{R}^{n_v}$ is the speed at the end of the time step $h$,
		\item unknown $\vect{\gamma}_\epsilon \in \mathbb{R}^{n_c}$ is the reaction in contacts and bilateral joints, a vector-signed measure with Lebesgue decomposition in atomic parts (impacts) and continuous parts (continuous reactions),
    \item $H \in \mathbb{R}^{n_v \times n_v}$ is a positive definite matrix containing $M$, 
		the block-diagonal matrix of the masses and inertia tensors of the bodies; if stiff
		loads are added too (ex. when finite elements come into play) it becomes a block-sparse matrix including the tangent stiffness and damping matrices,
		for instance $H=M-{h^2}\nabla_q\vect{f}-{h}\nabla_v\vect{f}$ in case of the first-order time steppers in \cite{TasoraAnitescuCMAME10,Francu2017},
    \item $D_\epsilon \in \mathbb{R}^{n_v \times n_c}$ is a sparse matrix, the transpose jacobian of all constraints,
    \item $\vect{k} \in \mathbb{R}^{n_v}$ is a vector containing terms proportional to applied forces $\vect{f}$ and to the last known speed,
    \item $\vect{b} \in \mathbb{R}^{n_c}$ is a vector containing constraint stabilization terms and rheonomic terms,
    \item $\Upsilon$ is the Cartesian product of all cones of admissible constraint forces, $\Upsilon = \bigtimes_j \Upsilon_j$, in detail:
    \item if a frictional contact is added, $\Upsilon_j \subset \mathbb{R}^3$ is a second order Lorentz cone with aperture proportional to friction coefficient,
    \item if a bilateral constraint is added, $\Upsilon_j = \mathbb{R}$ and $\Upsilon^\circ_j = \{0\}$,
    \item if a unilateral constraint is added, $\Upsilon_j = \mathbb{R}^+$ and $\Upsilon^\circ_j = \mathbb{R}^-$,
    \item $\Upsilon^{\circ}$ is the polar cone, opposite of the dual cone, i.e. $\Upsilon^{*} = -\Upsilon^\circ$.
\end{itemize}

@@

This CCP corresponds exactly to a first order optimality condition of a convex quadratic program:
\begin{subequations}
	\begin{empheq}[box=\fbox]{align}
	\text{min} \quad & \frac{1}{2} \vect{\gamma}_\epsilon^T N \vect{\gamma}_\epsilon + \vect{r}^T_\epsilon \vect{\gamma}_\epsilon \\
	\text{s.t.} \quad & \vect{\gamma}_\epsilon \in \Upsilon
	\end{empheq}
	\label{eq:ChronoCCP_min}
\end{subequations}
%
and in fact, the CCP \eqref{eq:ChronoCCP} can be written in the more conventional language of the Karush–Kuhn–Tucker optimality conditions on \textit{dual variables} $\vect{y}$ multipliers, for $\vect{y}=-\vect{u}_\epsilon$, and \textit{primal variables} $\vect{\gamma}_\epsilon$:
\begin{subequations}
	\begin{align}
    N \vect{\gamma}_\epsilon + \vect{r}_\epsilon + I \vect{y} &= \vect{0} \\
    \vect{\gamma}_\epsilon &= \vect{z} \\
    \Upsilon \ni \vect{z}  \quad \bot &\quad \vect{y} \in \Upsilon^\circ  
	\end{align}
	\label{eq:ChronoCCP_kkt}
\end{subequations}
After the convex program \eqref{eq:ChronoCCP_min} is solved by ADMM, one can compute 
\begin{align}
\vect{v} = H^{-1}( \vect{k} + D_\epsilon \vect{\gamma}_\epsilon)
\label{eq:v_post}
\end{align}
with an immediate step.

@@

This is an unconstrained convex quadratic program whose optimality conditions lead to the following linear problem:
\begin{align}
    \left[ N + \rho I \right] \vect{\gamma}_\epsilon^{k+1} =  \rho \vect{z}^k - \vect{r} - \vect{y}^k
		\label{eq:admm_1b}
\end{align}

@@

We experimented with the idea of using a $\rho$ value that changes on a per-constraint basis. If one could know in advance which contact will be active at the solution, one could split $\rho_i$ values between very high or very low values for the fastest convergence, however this is not possible because the set of active constraints is known only at the solution. However, we still can use a
simple heuristics consisting in forcing a small value $\rho_b = \SI{1e-9}{}$ for bilateral constraints, where one knows in advance that the corresponding dual variable $y_i$ would be null anyway. This produces better convergence results in systems containing both contacts and many bilateral joints, such as when simulating articulated robots that grasp some objects.
